---
title: "Storm Data Analysis"
author: "Anning Chen"
date: "January 23, 2016"
output: html_document
---

# Storm Data Analysis between 1950 and 2011


## Synopsis 

This analysis will look at storm data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database[https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2] across the United State from 1950 to 2011 and determine the type of events that  

1. are most harmful with respect to population health  
2. caused most economic damage

From the analysis, we found that 

1. tornados cause the most fatalities and injuries by far
2. flood causes the most amount of property damage and drought causes the most crop damage

See below for more detailed analysis and breakdown.

## Loading and Processing the Raw Data

Read in the unzipped repdata-data-StormData.csv file:
```{r cache = TRUE}
stormData <- read.csv('repdata-data-StormData.csv')
dim(stormData)
```
There are 902,297 rows and 37 columns in this dataset

Here are the first few lines of the data
```{r}
head(stormData)
```

## Fatalities and Injuries Analysis

There are 2 columns in the stormData dataframe that show impact to population health, namely: fatalities and injuries.  We will analyze these columns.

Sum up all the fatalities, get rid of the events with 0 fatalities and sort the list in descending order
```{r}
fatalities <- aggregate(x = stormData$FATALITIES, by = list(stormData$EVTYPE), sum)
fatalities <- subset(fatalities,x > 0)
fatalities <- fatalities[order(fatalities$x,decreasing = T),]
```

Using the 80-20 rule, we want to see the events that caused 80% of all fatalities and injuries

80% of the total number of fatalities is 12116
```{r}
.8*sum(fatalities$x)
```

So the 1st 11 events caused more than 80% of fatalities
```{r}
sum(head(fatalities$x,11))
```

Here are the 11 events that caused the most fatalities
```{r}
topFatalities <- head(fatalities,11)
topFatalities
```

Do the same with injuries
```{r}
injuries <- aggregate(x = stormData$INJURIES, by = list(stormData$EVTYPE), sum)
injuries <- subset(injuries,x > 0)
injuries <- injuries[order(injuries$x,decreasing = T),]
```

80% of the total number of injuries is 112423
```{r}
.8*sum(injuries$x)
```

The 1st 5 events caued more than 80% of injuries
```{r}
sum(head(injuries$x,5))
``` 

But for consistency, just show the same number of events as fatalities
```{r}
topInjuries <- head(injuries,11)
topInjuries
```

Plot of events that caused more than 80% of fatalities and injuries
```{r}
par(mfrow=c(1,2))
par(mar=c(7,4,4,1))
barplot(topFatalities$x,names.arg = topFatalities$Group.1, las = 3, cex.names = .75, col="red", ylab = 'Fatalities')
barplot(topInjuries$x,names.arg = topInjuries$Group.1, las = 3, cex.names = .75, col="orange", ylab='Injuries')
```

## Fatalities and Injuries Results

We can see from the plot above that tornados cause the most fatalities and injuries by far.

## Economic Consequences Analysis

There are 2 columns in the stormData dataFrame that show economic impact, namely: propdmg and cropdmg along with 2 other columns with exp as suffix used for units.  The analysis here is essentially the same as for fatalities and injuries but with an additional dollar conversion step.

Here is the propDmgConversion function handling the propDmgExp factors, any exp factor not recognized in the function will cause the value to be 0
```{r}
unique(stormData$PROPDMGEXP)

propDmgConversion <- function(x) {
	if(x['PROPDMGEXP'] == 'K' || x['PROPDMGEXP'] == 'k') {
		multBy <- 1000
	} else if (x['PROPDMGEXP'] == 'M' || x['PROPDMGEXP'] == 'm') {
		multBy <- 1000000
	} else if (x['PROPDMGEXP'] == 'H' || x['PROPDMGEXP'] == 'h') {
		multBy <- 100
	} else if (x['PROPDMGEXP'] == 'B' || x['PROPDMGEXP'] == 'b') {
		multBy <- 1000000000
	} else if (x['PROPDMGEXP'] %in% factor(c(0:8))) {
		multBy <- 10^as.numeric(as.character(x['PROPDMGEXP']))
	} else if (x['PROPDMGEXP'] == '+' || x['PROPDMGEXP'] == '-' || x['PROPDMGEXP'] == '') {
		multBy <- 1
	} else {
		multBy <- 0
	}
	as.numeric(x['PROPDMG'])*multBy
}
```

Convert all propDmg values to a consistent unit and add it as a new column PROPDMGCONV to stormData dataset:
```{r cache = TRUE}
stormData$PROPDMGCONV <- apply(stormData,MARGIN = 1,FUN = propDmgConversion)
```

Do the same for cropDmg values
```{r}
unique(stormData$CROPDMGEXP)

cropDmgConversion <- function(x) {
	if(x['CROPDMGEXP'] == 'K' || x['CROPDMGEXP'] == 'k') {
		multBy <- 1000
	} else if (x['CROPDMGEXP'] == 'M' || x['CROPDMGEXP'] == 'm') {
		multBy <- 1000000
	} else if (x['CROPDMGEXP'] == 'H' || x['CROPDMGEXP'] == 'h') {
		multBy <- 100
	} else if (x['CROPDMGEXP'] == 'B' || x['CROPDMGEXP'] == 'b') {
		multBy <- 1000000000
	} else if (x['CROPDMGEXP'] %in% factor(c(0:8))) {
		multBy <- 10^as.numeric(as.character(x['CROPDMGEXP']))
	} else if (x['CROPDMGEXP'] == '+' || x['CROPDMGEXP'] == '-' || x['CROPDMGEXP'] == '') {
		multBy <- 1
	} else {
		multBy <- 0
	}
	as.numeric(x['CROPDMG'])*multBy
}
```

Convert all cropDmg values to a consistent unit and add it as a new column CROPDMGCONV to stormData dataset:
```{r cache = TRUE}
stormData$CROPDMGCONV <- apply(stormData,MARGIN = 1,FUN = cropDmgConversion)
```

sum up all the propDmgConv, get rid of the events with 0 and sort the list in descending order
```{r}
propDmgConv <- aggregate(x = stormData$PROPDMGCONV, by = list(stormData$EVTYPE), sum)
propDmgConv <- subset(propDmgConv,x > 0)
propDmgConv <- propDmgConv[order(propDmgConv$x,decreasing = T),]
```

See that the top 10 ten events cause more than 80% of all propDmg, so just look at these 10 events
```{r}
.8*sum(propDmgConv$x)
sum(head(propDmgConv$x,10))
topPropDmg <- head(propDmgConv,10)
```

sum up all the cropDmgConv, get rid of the events with 0 and sort the list in descending order
```{r}
cropDmgConv <- aggregate(x = stormData$CROPDMGCONV, by = list(stormData$EVTYPE), sum)
cropDmgConv <- subset(cropDmgConv,x > 0)
cropDmgConv <- cropDmgConv[order(cropDmgConv$x,decreasing = T),]
```

See that the top 10 ten events cause more than 80% of all cropDmg, so just look at these 10 events
```{r}
.8*sum(cropDmgConv$x)
sum(head(cropDmgConv$x,10))
topCropDmg <- head(cropDmgConv,10)
```

Plot of events that caused more than 80% of propDmg and cropDmg
```{r}
par(mfrow=c(1,2))
par(mar=c(7,4,4,1))
barplot(topPropDmg$x,names.arg = topPropDmg$Group.1, las = 3, cex.names = .75, col="red", ylab = 'property damage $')
barplot(topCropDmg$x,names.arg = topCropDmg$Group.1, las = 3, cex.names = .75, col="orange", ylab='crop damage $')
```

## Economic Consequences Results

We can see that flood causes the most amount of property damage and drought causes the most crop damage.

